---
title: "Analysis on Storm Dataset"
date: "October 8, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis 

*Questions*

Your data analysis must address the following questions:

*1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

*2. Across the United States, which types of events have the greatest economic consequences?

## Data Processing

a. Load the data
```{r cache = TRUE}

rm(list = ls())
#set working directory
setwd("D:/Coursera/DS/Reproducible/DDS_RepR_w4")
getwd()
if(!file.exists("data")) {
    dir.create("data")
}
#download and unzip, file handling

fileURL<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileURL, destfile="./StormData.csv.bz2")
dt<-read.csv(paste(getwd(),"StormData.csv.bz2", sep="/"), stringsAsFactors = FALSE)
```
b. format and rename for convinence 
From the names of column, we're interested in the crop and population damages, so we only subsect those data. Here I use `select()` and `filter()` to subsect the dataframe. Alternative could use `subset()`. Also, `%>%`is used as pipe coding style. 

The list of them:

* `EVTYPE`: the event type,
* `BGN_DATE`: the date of the event,
* `FATALITIES`: the number of fatalities that resulted,
* `INJURIES`: the number of injuries that resulted,
* `PROPDMG`: an estimate for the amount of property damage,
* `PROPDMGEXP`: a "multiplier" extension for the amount in PROPDMG,
* `CROPDMG`: an estimate for the amount of crop damage,
* `CROPDMGEXP`: a "multiplier" extension for the amount in CROPDMG,
* `REMARKS`: remarks added by the person recording the data.

```{r echo=TRUE, warning=FALSE, cache=TRUE}
names(dt)
library("dplyr")
#only choose the above columns 
dt_sel <- dt %>%
    select(EVTYPE, BGN_DATE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP,
           CROPDMG, CROPDMGEXP, REMARKS)
#format the date by as.Date()
date_f<- dt_sel$BGN_DATE %>%
    strsplit(" ") %>%   #separate by the " ": first element Date, second element time
    unlist() %>%        #strsplit output list
    matrix(nrow=dim(dt_sel)[1], byrow=TRUE)
date_f<-as.Date(date_f[,1],format="%m/%d/%Y") #format the dates
dt_sel$DATE<-date_f 
dt_sel<-select(dt_sel,-BGN_DATE)  #remove the BGN_DATE, alternative is to overwrite this at the previous step
```

Let's look at the following variables:

* `PROPDMG`: an estimate for the amount of property damage,
* `PROPDMGEXP`: a "multiplier" extension for the amount in PROPDMG,
* `CROPDMG`: an estimate for the amount of crop damage,
* `CROPDMGEXP`: a "multiplier" extension for the amount in CROPDMG

```{r}
unique(dt_sel$PROPDMGEXP)
unique(dt_sel$CROPDMGEXP)
```
Both the `*EXP` has different characters standing for different multipliers.
Here we interprate the symbols as followings:

`H/h` as 10<sup>2</sup>
`K/k` as 10<sup>3</sup>
`M/m` as 10<sup>6</sup>
`B/b` as 10<sup>9</sup>

Here, we use `switch()` function to translate the characters into numerical variables.
```{r cache=TRUE}
#use sapply to loop around to change the characters into factors
prop_factor <- sapply(dt_sel$PROPDMGEXP,
    function(x) {switch(x,  "k" = 1000, "K" = 1000,
                            "m" = 1000000, "M" = 1000000, 
                            "B" = 1000000000, "b"= 1000000000,
                            "H"= 100, "h"=100, 1)})
#compare the factor to verify if the translation is correct
table(prop_factor)
table(dt_sel$PROPDMGEXP)

crop_factor <- sapply(dt_sel$CROPDMGEXP,
    function(x) {switch(x,  "k" = 1000, "K" = 1000,
                            "m" = 1000000, "M" = 1000000, 
                            "B" = 1000000000, "b"= 1000000000,
                            "H"= 100, "h"=100, 1)})
table(crop_factor)
table(dt_sel$CROPDMGEXP)
#Multiply the numbers by the factors
dt_sel_f<-dt_sel %>%
    mutate(CROP_VAL=CROPDMG*crop_factor,PROP_VAL=PROPDMG*prop_factor )
```

Now, we finish the manipulation of the data, and begin to answer the two questions. 

## Results

**1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?**

```{r cache=TRUE}
#change all the events to upper case
dt_sel_f$EVTYPE<-toupper(dt_sel_f$EVTYPE)
dt_sel_f %>%
    filter(EVTYPE!="?") %>% # filter out one event called "?"
    group_by(EVTYPE) %>%
    summarise(Death=sum(FATALITIES), Injury=sum(INJURIES), Property_value=sum(PROP_VAL), Crop_Value=sum(CROP_VAL))
```

*one should be able to futher compact the list as a few events have similar names and referring to the same events*

One can further examine the top ten events in each category
```{r cache=TRUE}
dt_sel_f %>%
    select(EVTYPE, FATALITIES) %>%
    group_by(EVTYPE) %>%
    summarise(Death=sum(FATALITIES)) %>%
    arrange(desc(Death)) %>%
    head(10) 
```

From this list, one can see **TORNADO** is the most fatal event.

```{r cache=TRUE}
dt_sel_f %>%
    select(EVTYPE, INJURIES) %>%
    group_by(EVTYPE) %>%
    summarise(Injury=sum(INJURIES)) %>%
    arrange(desc(Injury)) %>%
    head(10) 
```
From this list, one can see **TORNADO** is the event with most injuries.

Let's do a plot for this question. 

```{r cache=TRUE}

dt_sel_f$YEAR<-as.character(format(dt_sel_f$DATE, "%Y"))
tornadoes <- dt_sel_f %>%
    filter(EVTYPE == "TORNADO") %>%
    group_by(YEAR) %>%
    summarize(Death = sum(FATALITIES), Injury = sum(INJURIES))

par(mfrow=c(1,2))
plot(tornadoes$YEAR,tornadoes$Death, xlab = "Year", ylab = "Death", main = "Tornado Death plot")
plot(tornadoes$YEAR,tornadoes$Injury,xlab = "Year", ylab = "Injury", main = "Tornado Injury plot")
```

**2. Across the United States, which types of events have the greatest economic consequences?**

Similar to the above plots, we can generate the table for CROP and PROPERTY DAMAGE VALUES
```{r}
#Property_value=sum(PROP_VAL), Crop_Value=sum(CROP_VAL)
dt_sel_f %>%
    select(EVTYPE, PROP_VAL) %>%
    group_by(EVTYPE) %>%
    summarise(Property_value=sum(PROP_VAL)) %>%
    arrange(desc(Property_value)) %>%
    head(10) 

dt_sel_f %>%
    select(EVTYPE, CROP_VAL) %>%
    group_by(EVTYPE) %>%
    summarise(Crop_Value=sum(CROP_VAL)) %>%
    arrange(desc(Crop_Value)) %>%
    head(10) 

```

In term of property value, FLOOD is the most costly one. 
In term of the crops values, DROUGH is the one.

## Data source

https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2


